The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles

### 2022-11-08 ms/ms.Rmd
<edit ms>

---
title: |
  The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles 
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: R. Alan Covey
    affil: a
    email: R.Alan.Covey@austin.utexas.edu
  - name: Robert Z. Selden, Jr.
    affil: b
    email: zselden@sfasu.edu
  - name: Astrid Runggaldier
    affil: c
    email: astrid@austin.utexas.edu
  - name: Nicole D. Payntar
    affil: a
    email: npayntar@utexas.edu
affiliation:
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
  - num: b
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University 
  - num: c
    address: |
      Department of Art and Art History, The University of Texas at Austin; The Mesoamerica Center, The University of Texas at Austin
  - num: d
    address: |
      Department of Anthropology, The University of Texas at Austin
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  After centuries of looting along Peru’s North Coast, archaeologists acknowledge that the pottery of the Chimú Empire is one of the most collected Andean artifacts, but also one of the most poorly understood.  Much of the enduring classificatory uncertainty comes from the problematic provenance of most Chimú vessels, and the fact that the distinctive blackware identified as Chimú represents the production of workshops from across an extensive area, during periods of regional political decentralization (c. 900-1200 CE), imperial growth (c. 1200-1450), and foreign conquest by the Inca (c. 1450-1535) and Spanish (after 1532) empires.  This chapter builds on previous seriations and field observations, using geometric morphometric analysis of a sample of 3D-scanned Chimú bottles from publicly held collections at the University of Texas (Austin) and the American Museum of Natural History.  Since Chimú bottles were formed in workshops using molds, variations in vessel shape can serve as indicators of variable practices.  We compare a sample of common Chimú blackware bottles with a sample of _Inca-Chimú_ vessels that carry features typical of the short Inca occupation of the North Coast.  Differences between the two samples offer new lines of evidence that can guide more precise classifications of these vessels, as well as new interpretations of the imperial history of Peru’s North Coast.
keywords: |
  Sections; lists; figures; tables; mathematics; fonts; references; appendices
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

>>>>

---
title: |
  The Changing Shape of Empire: A Morphological Study of Curated Chimú Jars and Bottles 
type: ORIGINAL RESEARCH ARTICLE
author:
  - name: R. Alan Covey
    affil: a
    email: R.Alan.Covey@austin.utexas.edu
  - name: Robert Z. Selden, Jr.
    affil: b
    email: zselden@sfasu.edu
  - name: Astrid Runggaldier
    affil: c
    email: astrid@austin.utexas.edu
  - name: Nicole D. Payntar
    affil: a
    email: npayntar@utexas.edu
affiliation:
  - num: a
    address: |
      Department of Anthropology, The University of Texas at Austin
  - num: b
    address: |
      Heritage Research Center, Stephen F. Austin State University; Department of Biology, Stephen F. Austin State University; and Cultural Heritage Department, Jean Monnet University 
  - num: c
    address: |
      Department of Art and Art History, The University of Texas at Austin; The Mesoamerica Center, The University of Texas at Austin
  - num: d
    address: |
      Department of Anthropology, The University of Texas at Austin
bibliography: interactcadsample.bib
appendix: appendix.tex
abstract: |
  After centuries of looting along Peru’s North Coast, archaeologists acknowledge that the pottery of the Chimú Empire is one of the most collected Andean artifacts, but also one of the most poorly understood.  Much of the enduring classificatory uncertainty comes from the problematic provenance of most Chimú vessels, and the fact that the distinctive blackware identified as Chimú represents the production of workshops from across an extensive area, during periods of regional political decentralization (CE 900-1200), imperial growth (CE 1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish (after CE 1532) empires.  This chapter builds on previous seriations and field observations, using geometric morphometric analysis of a sample of 3D-scanned Chimú bottles from publicly held collections at the University of Texas at Austin.  Since Chimú bottles were formed in workshops using molds, variations in vessel shape can serve as indicators of variable practices.  We compare a sample of common Chimú blackware bottles with a sample of _Inca-Chimú_ vessels that carry features typical of the short Inca occupation of the North Coast.  Differences between the two vessel forms offer new lines of evidence that can guide more precise classifications, as well as new interpretations of the imperial history of Peru’s North Coast.
keywords: |
  Sections; lists; figures; tables; mathematics; fonts; references; appendices
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \def\tightlist{}
output: rticles::tf_article
---

### 2022-11-08 ms/ms.tex
<edit ms>

\begin{abstract}
After centuries of looting along Peru's North Coast, archaeologists
acknowledge that the pottery of the Chimú Empire is one of the most
collected Andean artifacts, but also one of the most poorly understood.
Much of the enduring classificatory uncertainty comes from the
problematic provenance of most Chimú vessels, and the fact that the
distinctive blackware identified as Chimú represents the production of
workshops from across an extensive area, during periods of regional
political decentralization (c.~900-1200 CE), imperial growth
(c.~1200-1450), and foreign conquest by the Inca (c.~1450-1535) and
Spanish (after 1532) empires. This chapter builds on previous seriations
and field observations, using geometric morphometric analysis of a
sample of 3D-scanned Chimú bottles from publicly held collections at the
University of Texas (Austin) and the American Museum of Natural History.
Since Chimú bottles were formed in workshops using molds, variations in
vessel shape can serve as indicators of variable practices. We compare a
sample of common Chimú blackware bottles with a sample of
\emph{Inca-Chimú} vessels that carry features typical of the short Inca
occupation of the North Coast. Differences between the two samples offer
new lines of evidence that can guide more precise classifications of
these vessels, as well as new interpretations of the imperial history of
Peru's North Coast.
\end{abstract}

>>>>

\begin{abstract}
After centuries of looting along Peru's North Coast, archaeologists
acknowledge that the pottery of the Chimú Empire is one of the most
collected Andean artifacts, but also one of the most poorly understood.
Much of the enduring classificatory uncertainty comes from the
problematic provenance of most Chimú vessels, and the fact that the
distinctive blackware identified as Chimú represents the production of
workshops from across an extensive area, during periods of regional
political decentralization (CE 900-1200), imperial growth (CE
1200-1450), and foreign conquest by the Inca (CE 1450-1535) and Spanish
(after CE 1532) empires. This chapter builds on previous seriations and
field observations, using geometric morphometric analysis of a sample of
3D-scanned Chimú bottles from publicly held collections at the
University of Texas at Austin. Since Chimú bottles were formed in
workshops using molds, variations in vessel shape can serve as
indicators of variable practices. We compare a sample of common Chimú
blackware bottles with a sample of \emph{Inca-Chimú} vessels that carry
features typical of the short Inca occupation of the North Coast.
Differences between the two vessel forms offer new lines of evidence
that can guide more precise classifications, as well as new
interpretations of the imperial history of Peru's North Coast.
\end{abstract}
